---
title: "DATA2050_WRU"
author: "Yixun Kang"
date: "2025-03-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install packages

```{r, message=FALSE}
library(devtools)
library(wru)
#library(zipWRUext2)
#devtools::install_github("https://github.com/jcuriel-unc/zipWRUext", subdir = "zipWRUext2")
library(MLmetrics)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
```

## Data Preprocessing

```{r}
# Remove surname = CRICKENBERGER and zcta = 27877
# Remove zcta in 23845, 23917, 24148, 30546, 37821
# Remove 46 rows with NA in BIFSG
data = read.csv("data.csv") # this is the cleaned/filtered data with 6975377 obs
# Map party
small_parties = c("NLB", "GRE", "WTP", "CST", "JFA", "UNA")
data$party_cd = ifelse(data$party_cd %in% small_parties, "IND", data$party_cd)
```

## WRU: BISG, fBISG, and BIFSG

```{r}
# BISG
Sys.setenv(CENSUS_API_KEY = "f22603397c6685d1c71d15779b942c97c0624004")
wru_bisg = wru::predict_race(data, 
                             census.surname = TRUE, 
                             census.geo = "zcta", 
                             census.key = Sys.getenv("CENSUS_API_KEY"),
                             census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                           states = "NC", 
                                                           sex = FALSE,
                                                           year = "2020",
                                                           census.geo = "zcta",
                                                           retry = 3,
                                                           county.list = NULL),
                             year = "2020",
                             retry = 3,
                             skip_bad_geos = TRUE,
                             model = "BISG",
                             impute.missing = FALSE)
```

```{r}
# fBISG
wru_fbisg = wru::predict_race(data, 
                              census.surname = TRUE, 
                              census.geo = "zcta", 
                              census.key = Sys.getenv("CENSUS_API_KEY"),
                              census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                            states = "NC", 
                                                            sex = FALSE,
                                                            year = "2020",
                                                            census.geo = "zcta",
                                                            retry = 3,
                                                            county.list = NULL),
                              year = "2020",
                              retry = 3,
                              skip_bad_geos = TRUE,
                              model = "fBISG",
                              impute.missing = FALSE,
                              control = list(seed = 123))
```

```{r}
# BIFSG
wru_bifsg = wru::predict_race(data, 
                              census.surname = TRUE, 
                              census.geo = "zcta", 
                              census.key = Sys.getenv("CENSUS_API_KEY"),
                              census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                            states = "NC", 
                                                            sex = FALSE,
                                                            year = "2020",
                                                            census.geo = "zcta",
                                                            retry = 3,
                                                            county.list = NULL),
                              year = "2020",
                              retry = 3,
                              skip_bad_geos = TRUE,
                              model = "BISG",
                              impute.missing = FALSE,
                              names.to.use = "surname, first")
```

## Weighted Estimator

```{r}
weighted_estimator = function(df, outcome, outcome_value, race_col = NULL, races = NULL, type) {
  outcome_mask = df[[outcome]] == outcome_value
  if (type == "pred") {
    df_race = df[, races]
    df_outcome = df_race[outcome_mask, ]
    n = colSums(df_outcome)
    d = colSums(df_race)
    we = n / d
  } else if (type == "actual") {
    df_outcome = df[outcome_mask, ]
    n = table(df_outcome[[race_col]])
    d = table(df[[race_col]])
    we = n / d
    race_order = c("white", "black", "hispanic", "api", "other")
    we = as.numeric(we[race_order])
    names(we) = race_order
  } else {
    stop("Must have a type")
  }
  return(we)
}
```

```{r}
# Predicted weighted estimator
# Using BISG result as an example
outcome_values = c("DEM", "REP", "LIB", "IND")
races = c("pred.whi", "pred.bla", "pred.his", "pred.asi", "pred.oth")
matrix = matrix(NA, nrow = length(races), ncol = length(outcome_values))
rownames(matrix) = races
colnames(matrix) = outcome_values
for (i in seq_along(outcome_values)) {
  party = outcome_values[i]
  we = weighted_estimator(wru_bisg, "party_cd", party, NULL, races, "pred")
  matrix[, party] = round(we, 5)
}
pred_wes = as.data.frame(matrix)
#write.csv(pred_wes, "file_name", row.names = TRUE)
```

```{r}
# Actual weighted estimator
matrix = matrix(NA, nrow = 5, ncol = 4)
rownames(matrix) = races
colnames(matrix) = outcome_values
for (i in seq_along(outcome_values)) {
  party = outcome_values[i]
  we = weighted_estimator(data, "party_cd", party, "true_race", NULL, "actual")
  matrix[, party] = round(we, 5)
}
actual_wes = as.data.frame(matrix)
#write.csv(actual_wes, "file_name", row.names = TRUE)
```

## Surname Perturbation on 5%, 10% and 20% of the data

- Can also do perturbation combinations like 5% surname perturbation + 5% ZCTA perturbation
- Use the code in python to export the ZCTA perturbed data sets and change the data used in `data_xyz = data %>% mutate(row_id = row_number())`, for example, `data_xyz = data_20 %>% mutate(row_id = row_number())`, where data_20 is the exported data set with 20% ZCTA perturbation
- NOTE: if using ZCTA perturbed data sets like data_20, do this first:
  - `data_20 = read.csv("data_zp_20.csv")`
  - `data_20 = data_20 %>% select(-zcta) %>% rename(zcta = Swapped_zcta)`

```{r}
set.seed(123)

# Change data to ZCTA perturbed data sets if doing perturbation combinations 
# Below is an example of surname perturbation only
data_xyz = data %>% mutate(row_id = row_number())

n_total = nrow(data_xyz)

# Change the perturbation percentage to 0.05 or 0.2
n_select = floor(0.1 * n_total)

select_id = sample(data_xyz$row_id, n_select)
data_xyz$surname = ifelse(data_xyz$row_id %in% select_id,
                          paste0(data_xyz$surname, "xyz"),
                          data_xyz$surname)
```

```{r}
# BISG
wru_bisg = wru::predict_race(data_xyz, 
                             census.surname = TRUE, 
                             census.geo = "zcta", 
                             census.key = Sys.getenv("CENSUS_API_KEY"),
                             census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                           states = "NC", 
                                                           sex = FALSE,
                                                           year = "2020",
                                                           census.geo = "zcta",
                                                           retry = 3,
                                                           county.list = NULL),
                             year = "2020",
                             retry = 3,
                             skip_bad_geos = TRUE,
                             model = "BISG",
                             impute.missing = FALSE)
wru_bisg = wru_bisg %>% filter(!is.na(pred.whi))
#write.csv(wru_bisg, "file_name", row.names = FALSE)
```

```{r}
# ZCTA perturbation will result missing ZCTA, filter out these rows
# Use data_zp20_sp10 in fBISG and BIFSG
# zp20 means 20% ZCTA perturbation and sp10 means 10% surname perturbation
# This is an example of passing the 20% ZCTA perturbed data set into the chunks above
data_zp20_sp10 = wru_bisg[, 1:(ncol(wru_bisg) - 5)] 
```

```{r}
# fBISG
# The input data set should be the filtered data set with no missing ZCTA
wru_fbisg = wru::predict_race(data_zp20_sp10, 
                              census.surname = TRUE, 
                              census.geo = "zcta", 
                              census.key = Sys.getenv("CENSUS_API_KEY"),
                              census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                            states = "NC", 
                                                            sex = FALSE,
                                                            year = "2020",
                                                            census.geo = "zcta",
                                                            retry = 3,
                                                            county.list = NULL),
                              year = "2020",
                              retry = 3,
                              skip_bad_geos = TRUE,
                              model = "fBISG",
                              impute.missing = FALSE,
                              control = list(seed = 123))
wru_fbisg = wru_fbisg %>% filter(!is.na(pred.whi))
#write.csv(wru_fbisg, "file_name", row.names = FALSE)
```  

```{r}
# BIFSG
# Use batch if memory is used up
n = nrow(data_zp20_sp10)
idx1 = 1:floor(n / 2)
idx2 = (floor(n / 2) + 1):n
data1 = data_zp20_sp10[idx1, ]
data2 = data_zp20_sp10[idx2, ]

wru_bifsg1 = wru::predict_race(data1, 
                               census.surname = TRUE, 
                               census.geo = "zcta", 
                               census.key = Sys.getenv("CENSUS_API_KEY"),
                               census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"), 
                                                             states = "NC", 
                                                             sex = FALSE,
                                                             year = "2020",
                                                             census.geo = "zcta",
                                                             retry = 3,
                                                             county.list = NULL),
                               year = "2020",
                               retry = 3,
                               skip_bad_geos = TRUE,
                               model = "BISG",
                               impute.missing = FALSE,
                               names.to.use = "surname, first")

wru_bifsg2 = wru::predict_race(data2, 
                               census.surname = TRUE, 
                               census.geo = "zcta", 
                               census.key = Sys.getenv("CENSUS_API_KEY"),
                               census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                             states = "NC", 
                                                             sex = FALSE,
                                                             year = "2020",
                                                             census.geo = "zcta",
                                                             retry = 3,
                                                             county.list = NULL),
                               year = "2020",
                               retry = 3,
                               skip_bad_geos = TRUE,
                               model = "BISG",
                               impute.missing = FALSE,
                               names.to.use = "surname, first")

wru_bifsg = rbind(wru_bifsg1, wru_bifsg2)
wru_bifsg = wru_bifsg %>% filter(!is.na(pred.whi))
#write.csv(wru_bifsg, "file_name", row.names = FALSE)
```

## ZCTA Perturbation

- This is only ZCTA perturbation
- If doing perturbation combinations, using the code above under Surname Perturbation

```{r}
# BISG
# An example of 20% ZCTA perturbation
data_20 = read.csv("data_zp_20.csv")
data_20 = data_20 %>% select(-zcta) %>% rename(zcta = Swapped_zcta)

wru_bisg = wru::predict_race(data_20, 
                             census.surname = TRUE, 
                             census.geo = "zcta", 
                             census.key = Sys.getenv("CENSUS_API_KEY"),
                             census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                           states = "NC", 
                                                           sex = FALSE,
                                                           year = "2020",
                                                           census.geo = "zcta",
                                                           retry = 3,
                                                           county.list = NULL),
                             year = "2020",
                             retry = 3,
                             skip_bad_geos = TRUE,
                             model = "BISG",
                             impute.missing = FALSE)
wru_bisg = wru_bisg %>% filter(!is.na(pred.whi))

# ZCTA perturbation will result missing ZCTA, filter out these rows
# Use data_20_new in fBISG and BIFSG
data_20_new = wru_bisg[, 1:(ncol(wru_bisg) - 5)]

#write.csv(wru_bisg, "file_name", row.names = TRUE)
```

```{r}
# fBISG
wru_fbisg = wru::predict_race(data_20_new, 
                              census.surname = TRUE, 
                              census.geo = "zcta", 
                              census.key = Sys.getenv("CENSUS_API_KEY"),
                              census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                            states = "NC", 
                                                            sex = FALSE,
                                                            year = "2020",
                                                            census.geo = "zcta",
                                                            retry = 3,
                                                            county.list = NULL),
                              year = "2020",
                              retry = 3,
                              skip_bad_geos = TRUE,
                              model = "fBISG",
                              impute.missing = FALSE,
                              control = list(seed = 123))
wru_fbisg = wru_fbisg %>% filter(!is.na(pred.whi))
#write.csv(wru_fbisg, "file_name", row.names = FALSE)
```

```{r}
# BIFSG
wru_bifsg = wru::predict_race(data_20_new, 
                              census.surname = TRUE, 
                              census.geo = "zcta", 
                              census.key = Sys.getenv("CENSUS_API_KEY"),
                              census.data = get_census_data(key = Sys.getenv("CENSUS_API_KEY"),
                                                            states = "NC", 
                                                            sex = FALSE,
                                                            year = "2020",
                                                            census.geo = "zcta",
                                                            retry = 3,
                                                            county.list = NULL),
                              year = "2020",
                              retry = 3,
                              skip_bad_geos = TRUE,
                              model = "BISG",
                              impute.missing = FALSE,
                              names.to.use = "surname, first")
wru_bifsg = wru_bifsg %>% filter(!is.na(pred.whi))
#write.csv(wru_bifsg, "file_name", row.names = FALSE)
```

## Weighted Estimator

- Same as above

```{r}
outcome_values = c("DEM", "REP", "LIB", "IND")
races = c("pred.whi", "pred.bla", "pred.his", "pred.asi", "pred.oth")
matrix = matrix(NA, nrow = length(races), ncol = length(outcome_values))
rownames(matrix) = races
colnames(matrix) = outcome_values
for (i in seq_along(outcome_values)) {
  party = outcome_values[i]
  we = weighted_estimator(wru_bifsg_sp, "party_cd", party, NULL, races, "pred")
  matrix[, party] = round(we, 5)
}
pred_wes = as.data.frame(matrix)
#write.csv(pred_wes, "file_name", row.names = TRUE)
```

```{r}
matrix = matrix(NA, nrow = 4, ncol = 4)
rownames(matrix) = c("pred.bla", "pred.his", "pred.asi", "pred.oth")
colnames(matrix) = outcome_values
for (party in outcome_values) {
  outcome_mask = data_xyz[["party_cd"]] == party
  df_outcome = data_xyz[outcome_mask, ]
  n = table(df_outcome[["true_race"]])
  d = table(data_xyz[["true_race"]])
  we = n / d
  race_order = c("black", "hispanic", "api", "other")
  we = as.numeric(we[race_order])
  names(we) = race_order
  matrix[, party] = round(we, 5)
}
actual_wes = as.data.frame(matrix)
#write.csv(actual_wes, "file_name", row.names = TRUE)
```




